---
title: "BestSurvCutPoints"
output: html_document
date: "2023-09-18"
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction

This document provides an overview of the `BestSurvCutPoints` R package
for survival analysis. It explains the package's functionality,
demonstrates its usage with examples, and offers insights into
interpreting the results.

## Package Description

The `BestSurvCutPoints` package is designed to help researchers and data
analysts find optimal cutpoints in survival analysis. It utilizes
various functions, such as `findcutnum`, `findcut`, `findnumCox`, and
`findcutCox`, to assist in categorizing continuous variables into
intervals for improved survival data analysis.

It involves categorizing a continuous variable into intervals to best
distinguish event occurrence (e.g., death or disease) from
non-occurrence. This package utilises AIC or cross-validation to find
the ideal number of cutpoints, and genetic algorithms for their precise
locations and help researchers and data analysts determine the most
suitable cutpoints for survival data analysis.

Why is finding optimal cutpoints important in survival analysis? How can
it benefit researchers and data analysts? Providing answers to these
questions can make your document more informative.

# Installation

To use the `BestSurvCutPoints` package, you can install it from GitHub
using the following code:

```{r setup, warning=FALSE, message=FALSE}
# Install the package
# devtools::install_github("paytonyau/BestSurvCutPoints")
library(BestSurvCutPoints)
```

Loading Libraries In addition to BestSurvCutPoints, several other
libraries are required for comprehensive survival analysis. These
include survival, KMsurv, xtable, splines, pROC, and aod. Each library
serves a specific purpose in the analysis.

```{r library, warning=FALSE, message=FALSE}
# Load the necessary libraries
library("survival")   # Survival analysis library
library("KMsurv")     # Kaplan-Meier survival curves
library("xtable")     # Table generation for reports
library("splines")    # Basis splines for modeling
library("pROC")       # Receiver Operating Characteristic (ROC) analysis
library("aod")        # Analysis of Overdispersed Data
```

Example 1: Survival Data This section demonstrates the use of the
BestSurvCutPoints package with an example dataset related to survival
analysis.

### Data Description

#### Survival data example 1

+--------------+----------------------------------------------------+
| Column       | Explanation                                        |
+==============+====================================================+
| BMI          | body mass index                                    |
+--------------+----------------------------------------------------+
| event        | censoring status\                                  |
|              | 0=censored\                                        |
|              | 1=dead                                             |
+--------------+----------------------------------------------------+
| OS           | time of observation in months\                     |
|              | (minimum of event time and right censoring time)   |
+--------------+----------------------------------------------------+
| invasion     | depth of tumour invasion into stroma               |
+--------------+----------------------------------------------------+
| LVSI         | lymphovascular space invasion\                     |
|              | 0 = not seen\                                      |
|              | 1 = present                                        |
+--------------+----------------------------------------------------+

```{r Example 1, warning=FALSE, message=FALSE}
BMI = c(30,16,29,29,21,29,27,24,17,27,22,27,26,16,21,
        21,23,20,25,23,28,20,22,22,37,23,25,34,31,26)

event = c(1,1,1,1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0,1,0,0,1,0,1,0,0,1,1,0)

OS = c(138,92,64,15,62,235,214,197,41,33,257,115,123,44,154,
       71,61,182,75,214,25,217,113,200,175,117,166,0,57,186)

invasion = c(1.1,0.1,1.0,0.8,1.2,1.0,0.3,1.0,0.4,0.6,0.4,0.8,1.0,1.0,1.1,
             0.5,0.9,1.0,0.6,0.1,1.1,0.4,0.4,1.1,1.1,0.4,1.1,1.2,0.9,0.8)

LVSI = c(0,0,0,1,0,1,0,0,1,1,0,1,1,1,0,0,0,1,1,1,1,0,1,0,0,0,0,1,0,1)

# Create a matrix
dat1 = as.data.frame(cbind(BMI, event, OS, invasion, LVSI))

rm(BMI, event, OS, invasion, LVSI)
```

#### Survival data example 2 (Lung data from survival package)

+--------------+-----------------------------------------------------+
| Column       | Explanation                                         |
+==============+=====================================================+
| inst         | Institution code                                    |
+--------------+-----------------------------------------------------+
| time         | Survival time in days                               |
+--------------+-----------------------------------------------------+
| status       | censoring status\                                   |
|              | 1=censored\                                         |
|              | 2=dead                                              |
+--------------+-----------------------------------------------------+
| age          | Age in years                                        |
+--------------+-----------------------------------------------------+
| sex          | 1 = Male\                                           |
|              | 2 = Female                                          |
+--------------+-----------------------------------------------------+
| ph.ecog      | ECOG performance score as rated by the physician\   |
|              | 0=asymptomatic\                                     |
|              | 1 = symptomatic but completely ambulatory\          |
|              | 2 = in bed \<50% of the day\                        |
|              | 3 = in bed \> 50% of the day but not bedbound\      |
|              | 4 = bedbound                                        |
+--------------+-----------------------------------------------------+
| ph.karno     | Karnofsky performance score rated by physician\     |
|              | 0 = bad\                                            |
|              | 100 = good                                          |
+--------------+-----------------------------------------------------+
| pat.karno    | Karnofsky performance score as rated by patient\    |
|              | 0 = bad\                                            |
|              | 100 = good                                          |
+--------------+-----------------------------------------------------+
| meal.cal     | Calories consumed at meals                          |
+--------------+-----------------------------------------------------+
| wt.loss      | Weight loss in last six months                      |
+--------------+-----------------------------------------------------+

```{r Example 2, warning=FALSE, message=FALSE}
# The example data the survival package - from Example 2

dat2 = lung

# remove NA data to simpl
dat2 <- na.omit(dat2)  

# Recode the 'status' column
dat2$status <- ifelse(dat2$status == 1, 0, 1)

# ph.ecog to ph.ecog2
dat2$ph.ecog2 <- as.numeric(ifelse(dat2$ph.ecog %in% c(0, 1), 0, 
                                  ifelse(dat2$ph.ecog %in% 2:5, 1, NA)))

# ph.karno + pat.karno / 2
dat2$avg_karno <- (dat2$ph.karno + dat2$pat.karno) / 2
```

## findcutnum

### Function to find optimal cutpoints based on BMI for survival analysis

##### Example data 1
```{r findcutnum 1, warning=FALSE, message=FALSE}
findcutnum(
  factor = dat1$BMI,            # Predictor variable
  outcome = cbind(dat1$event, 
                  dat1$OS),     # Survival outcome data
  datatype = "survival",        # Data type for analysis
  nmin = 5,                     # Minimum group size
  segment = 100)                # Number of segments)
```

##### Example data 2
```{r findcutnum 2, warning=FALSE, message=FALSE}
findcutnum(factor = dat2$meal.cal, 
           outcome = cbind(dat2$status, 
                           dat2$time),
           datatype = "survival", 
           nmin = 5, 
           segment = 50)
```

### Function to find optimal cutpoints based on invasion for logistic regression analysis
##### Example 1
```{r findcutnum 1, warning=FALSE, message=FALSE}
findcutnum(
  factor = dat1$invasion,  # Predictor variable (invasion)
  outcome = dat1$LVSI,     # Outcome variable (LVSI)
  datatype = "logistic",   # Data type for logistic regression
  nmin = 5,                # Minimum group size
  segment = 100)           # Number of segments
```
##### Example 2
```{r findcutnum 2, warning=FALSE, message=FALSE}
findcutnum(
  factor = dat2$avg_karno,  # Predictor variable (ph.karno + pat.karno / 2)
  outcome = dat2$ph.ecog2,  # Outcome variable (ph.ecog2)
  datatype = "logistic",    # Data type for logistic regression
  nmin = 5,                 # Minimum group size
  segment = 25)            # Number of segments
```

## findcut

### Function to find optimal cutpoints based on BMI for survival analysis with 2 cutpoints

##### Example 1
```{r findcut_1, warning=FALSE, message=FALSE}
findcut(
  factor = dat1$BMI,                   # Predictor variable
  outcome = cbind(dat1$event, dat1$OS), # Survival outcome data
  cutnum = 2,                         # Number of cutpoints
  datatype = "survival",              # Data type for analysis
  nmin = 5,                           # Minimum group size
  segment = 100)                      # Number of segments
```

##### Example 2
```{r findcut_2, warning=FALSE, message=FALSE}
findcut(
  factor = dat2$meal.cal,                    # Predictor variable 
  outcome = cbind(dat2$status, dat2$time),   # Survival outcome data
  cutnum = 2,                                # Number of cutpoints
  datatype = "survival",                     # Data type for analysis
  nmin = 2,                                  # Minimum group size
  segment = 25)                             # Number of segments
```

### Function to find optimal cutpoints based on invasion for logistic regression analysis with 2 cutpoints

##### Example 1 
```{r findcut 1, warning=FALSE, message=FALSE}
findcut(
  factor = dat1$invasion,   # Predictor variable (invasion)
  outcome = dat1$LVSI,      # Outcome variable (LVSI)
  cutnum = 3,              # Number of cutpoints
  datatype = "logistic",   # Data type for logistic regression
  nmin = 5,                # Minimum group size
  segment = 100            # Number of segments
)
```

##### Example 2 

Error message appeared: `Error in qr.solve(L %*% V %*% t(L)) : singular matrix 'a' in solve`
meaning that the cutnum of the assigned number does not allow you to continue the run. Try lower the number of the `cutsum` argument
```{r findcut 2, warning=FALSE, message=FALSE}
findcut(
  factor = dat2$avg_karno,  # Predictor variable (ph.karno + pat.karno / 2)
  outcome = dat2$ph.ecog2,  # Outcome variable (ph.ecog2)
  cutnum = 2,              # Number of cutpoints  
  datatype = "logistic",    # Data type for logistic regression
  nmin = 5,                 # Minimum group size
  segment = 10)            # Number of segments
```

### The `findnumCox` finds the optimal number of cutpoints for survival analysis

```{r findcut, warning=FALSE, message=FALSE}
# Load your dataset from a CSV file (toydata.csv should be in the current working directory)
data <- read.csv("toydata.csv")

# Attach the data for easy access to columns
attach(data)

# Set a random seed for reproducibility
set.seed(30)

# Record the starting time
ptm <- proc.time()

# Call the findnumCox function to find optimal cutpoints

result <- findnumCox(
  target = BMI,             # Predictor variable (BMI)
  event = Death,            # Event indicator variable (Death)
  time = Death_surtime,     # Survival time variable (Death_surtime)
  confound = stage3,        # Confounding variable (stage3)
  totalcut = 3,             # Total number of cutpoints to consider
  initial_rr = NULL,        # Initial relative risk values (set to NULL)
  initial_cut = NULL,       # Initial cutpoint values (set to NULL)
  initial_domain = NULL,    # Initial domain values (set to NULL)  
  numgen = 10,              # Number of generations for genetic algorithm
  numcross = 20,            # Number of cross-validations
  gap = NULL                # Gap value (set to NULL)
)

# Calculate the time taken for the analysis
proc.time() - ptm

detach(data)
# Print the corrected AIC values
result$aic

# Find the number of optimal cutpoints (the index with the minimum AIC)
which.min(result$aic)

# Print the corrected hazard ratios (or relative risks)
result$HR

# Note: Corrected p-values for each coefficient estimator (Cutpvalue) are available in the result but are not printed here.
```

```{r findnumCox, warning=FALSE, message=FALSE}
# Attach the data for easy access to columns
attach(dat1)

# Set a random seed for reproducibility
set.seed(30)

# Record the starting time
ptm <- proc.time()

# Call the findnumCox function to find optimal cutpoints

result <- findnumCox(
  target = BMI,             # Predictor variable (BMI)
  event = event,            # Event indicator variable (Death)
  time = OS,                # Survival time variable (Death_surtime)
  confound = LVSI,          # Confounding variable (stage3)
  totalcut = 2,             # Total number of cutpoints to consider
  initial_rr = NULL,        # Initial relative risk values (set to NULL)
  initial_cut = NULL,       # Initial cutpoint values (set to NULL)
  initial_domain = NULL,    # Initial domain values (set to NULL)  
  numgen = 10,              # Number of generations for genetic algorithm
  numcross = 20,            # Number of cross-validations
  gap = NULL                # Gap value (set to NULL)
)

# Calculate the time taken for the analysis
proc.time() - ptm

detach(dat1)
# Print the corrected AIC values
result$aic

# Find the number of optimal cutpoints (the index with the minimum AIC)
which.min(result$aic)

# Print the corrected hazard ratios (or relative risks)
result$HR

# Note: Corrected p-values for each coefficient estimator (Cutpvalue) are available in the result but are not printed here.
```



```{r findnumCox, warning=FALSE, message=FALSE}
attach(dat2)

result <- findnumCox(
  target = avg_karno,       # Predictor variable (BMI)
  event = status,           # Event indicator variable (Death)
  time = time,              # Survival time variable (Death_surtime)
  confound = ph.ecog2,      # Confounding variable (stage3)
  totalcut = 2,             # Total number of cutpoints to consider
  initial_rr = NULL,        # Initial relative risk values (set to NULL)
  initial_cut = NULL,       # Initial cutpoint values (set to NULL)
  initial_domain = NULL,    # Initial domain values (set to NULL)  
  numgen = 10,              # Number of generations for genetic algorithm
  numcross = 20,            # Number of cross-validations
  gap = NULL                # Gap value (set to NULL)
)
```

#### 
##### confound example
```{r findcutCox, warning=FALSE, message=FALSE}
# Set a random seed for reproducibility
set.seed(2019)

# Extract data from your dataset
target <- data$BMI
event <- data$Death
time <- data$Death_surtime
confound <- data$stage3

# Create user data based on input and handle missing values
userdata <- na.omit(data.frame(target, event, time, confound))

# Determine the number of observations
N <- dim(userdata)[1]

# Generate random bootstrap samples
numboot <- sample(N)

# Define initial values for relative risks, cutpoints, and domains
initial_rr <- list(c(3, 2), c(3, 6, 2), c(3, 4, 5, 2))
initial_cut <- list(c(19), c(19, 30), c(15, 25, 30))
initial_domain <- list(
  matrix(c(12, 35, 0, 5, 0, 5), ncol = 2, byrow = TRUE),
  matrix(c(15, 35, 15, 35, 0, 5, 0, 5, 0, 5), ncol = 2, byrow = TRUE),
  matrix(c(15, 35, 15, 35, 15, 35, 0, 5, 0, 5, 0, 5, 0, 5), ncol = 2, byrow = TRUE)
)

# Example for no initial values provided
E <- findcutCox(BMI, Death, Death_surtime, stage3, numcut = 3, initial_rr = NULL,
                initial_cut = NULL, initial_domain = NULL, numgen = 15, gap = NULL)

# Print the result
E

# Example for providing initial values
G <- findcutCox(BMI, Death, Death_surtime, stage3, numcut = 3, initial_rr = initial_rr,
                initial_cut = initial_cut, initial_domain = initial_domain, numgen = 15, gap = NULL)

# Print the result
G
