---
title: "BestSurvivalCuts"
author: "Payton Yau"
date: "2023-09-18"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    number_sections: yes
abstract: This document provides an overview of the BestSurvivalCuts package.
keywords:
- R
- BestSurvivalCuts
- survival analysis
bibliography: bibliography.bib
csl: apa.csl
subtitle: An Introduction to BestSurvivalCuts
editor_options:
  markdown:
    wrap: 72
---

This document serves as a comprehensive guide to the `BestSurvivalCuts`
R package, a tool designed for survival analysis. It elucidates the
functionalities of the package, provides practical examples of its
application, and offers guidance on how to interpret the results
effectively.

### Package Description

The `BestSurvivalCuts` package aids in finding optimal cutpoints in
survival analysis. It uses functions like `findcutnum`, `findcut`,
`findnumCox`, and `findcutCox` to categorize continuous variables into
intervals.

This package using AIC or cross-validation, determines the ideal number
of cutpoints and employs genetic algorithms for their precise locations.
These cutpoints are instrumental in identifying significant thresholds
in predictors, which can indicate different risk levels or stages of a
disease. Not only do they aid in classifying patients into groups for
personalized care, but they also find application in plant science. They
can help pinpoint crucial thresholds in plant growth, disease
resistance, or other essential traits, thereby facilitating a more
nuanced understanding and care of different plant species.

### Installation

To use the `BestSurvivalCuts` package, you can install it from GitHub
using the following code:

```{r setup, warning=FALSE, message=FALSE}
# Install the package
# devtools::install_github("paytonyau/BestSurvivalCuts")
library("BestSurvivalCuts")
```

Loading Libraries required for `BestSurvivalCuts`, these include
`survival`, `KMsurv`, `xtable`, `splines`, `pROC`, and `aod`.

```{r library, warning=FALSE, message=FALSE}
# Load the necessary libraries
library("survival")   # Survival analysis library
library("KMsurv")     # Kaplan-Meier survival curves
library("xtable")     # Table generation for reports
library("splines")    # Basis splines for modeling
library("pROC")       # Receiver Operating Characteristic (ROC) analysis
library("aod")        # Analysis of Overdispersed Data
```

### Survival data

To illustrate how to use the package, the following data contained
information of 30 individuals:

```{r Example 1, warning=FALSE, message=FALSE}
BMI = c(30,16,29,29,21,29,27,24,17,27,22,27,26,16,21,
        21,23,20,25,23,28,20,22,22,37,23,25,34,31,26)

event = c(1,1,1,1,0,0,1,0,0,0,0,0,0,0,0,0,0,0,0,1,0,0,1,0,1,0,0,1,1,0)

OS = c(138,92,64,15,62,235,214,197,41,33,257,115,123,44,154,
       71,61,182,75,214,25,217,113,200,175,117,166,0,57,186)

invasion = c(1.1,0.1,1.0,0.8,1.2,1.0,0.3,1.0,0.4,0.6,0.4,0.8,1.0,1.0,1.1,
             0.5,0.9,1.0,0.6,0.1,1.1,0.4,0.4,1.1,1.1,0.4,1.1,1.2,0.9,0.8)

LVSI = c(0,0,0,1,0,1,0,0,1,1,0,1,1,1,0,0,0,1,1,1,1,0,1,0,0,0,0,1,0,1)

# Create a matrix
dat1 = as.data.frame(cbind(BMI, event, OS, invasion, LVSI))

rm(BMI, event, OS, invasion, LVSI)
```

+---------------+------------------------------------------+
| Column        | Explanation                              |
+===============+==========================================+
| BMI           | body mass index                          |
+---------------+------------------------------------------+
| event         | censoring status\                        |
|               | 0=censored\                              |
|               | 1=dead                                   |
+---------------+------------------------------------------+
| OS            | time of observation in months\           |
|               | (minimum of event time and right         |
|               | censoring time)                          |
+---------------+------------------------------------------+
| invasion      | depth of tumour invasion into stroma     |
+---------------+------------------------------------------+
| LVSI          | lymphovascular space invasion\           |
|               | 0 = not seen\                            |
|               | 1 = present                              |
+---------------+------------------------------------------+

### findcutnum

#### A function to find optimal cutpoints based on BMI for survival analysis

First use `findcutnum` to decide the optimal number of cutoff points
based on AIC values.

```{r findcutnum_1, warning=FALSE, message=FALSE}
findcutnum(
  # continuous risk factor which needs the optimal number of cutoffs
  factor = dat1$BMI,
  # (event, time),
  # event(1:event occurs;0:right censoring),
  # time(minimum of event time and right censoring time),so outcome dimension is N by 2
  outcome = cbind(dat1$event, dat1$OS),
  # the number of cutoffs
  cutnum=NA,
  # Data type for analysis
  datatype = "survival",
  # the minimum number of each group and the default is 20
  nmin = 8,
  # the total number of pieces and the default is 100
  segment = 100)
```

In the dataset containing only 30 entries, the minimum number of each
segment is set to 8 (nmin=8), while the default number of segments is
maintained at 100 (segment=100). The goal is to find the optimal number
of cutoff points for BMI in relation to the event of death. Thus, BMI is
the variable factor, and the outcome is a 30 x 2 matrix generated with
the `cbind` function.

The AIC values are as follows: 50.52894 for no cutoff point, 49.20829
for one cutoff, and 51.10694 for two. Given the small size of the
sample, it isn't realistic to make 3 or more cutoff points, and the AIC
values for 3 and 4 cutoffs are both infinite.

The smallest AIC value is 49.20829, suggesting that one cutoff point
would be optimal for dividing these patients into low and high-risk
groups for death, assuming there are no clinical or other concerns.

### findcut

#### Function to find optimal cutpoints based on BMI for survival analysis with 2 cutpoints

Than use `findcut` to find the optimal location for the above BMI data.

```{r findcut_1, warning=FALSE, message=FALSE}
findcut(
    factor = dat1$BMI,                   # Predictor variable
    outcome = cbind(dat1$event, dat1$OS), # Survival outcome data
    cutnum = 2,                         # Number of cutpoints
    datatype = "survival",              # Data type for analysis
    nmin = 8,                           # Minimum group size
    segment = 100)                      # Number of segments

```

+---------------+-------------------------------------------------+
| Column        | Explaination                                    |
+===============+=================================================+
| allcut        | (only the first five results are shown here)    |
+---------------+-------------------------------------------------+
| Cut1 & Cut2   | locations of the two cutoff points              |
+---------------+-------------------------------------------------+
| Log.rank.test | when the cutoff points were placed at           |
|               | (Cut1,Cut2), log-rank test was used to test the |
|               | significance of the points, i.e., p-value of    |
|               | the log-rank test.                              |
+---------------+-------------------------------------------------+
| Likelihood.   | when the cutoff points were placed at           |
| ratio.test    | (Cut1,Cut2), likelihood ratio test was used to  |
|               | test the significance of the points, i.e.,      |
|               | p-value of the likelihood ratio test (log       |
|               | test).                                          |
+---------------+-------------------------------------------------+

In this dataset of 30 entries, the number of each segment is set to 8
(nmin=8), while the number of segments remains at the default value of
100 (segment=100). The goal is to find the optimal locations of cutoff
points for BMI in response to event death. Therefore, BMI is the
variable factor, and the outcome is a 30 x 2 matrix generated with the
`cbind` function. For illustrative purposes, it was arbitrarily decided
to create 3 groups of BMI, or 2 cutoff points.

The best locations of the cutoff points are determined by two methods:

-   `logranktest`: This method tests all possible cutoffs using the log
    rank test. The cutoff that yields the minimal p-value is considered
    the best location.

-   `logtest_likelihood.ratio.test`: This method tests all possible
    cutoffs using the likelihood ratio test. The cutoff that yields the
    minimal p-value is considered the best location.

## Dichotomised data

#### Function to find optimal cutpoints based on the invasion for logistic regression analysis with 2 cutpoints

Here, the function `findcutnum` is utilised to determine the optimal
number of cutoff points based on AIC values.

```{r findcutnum_2, warning=FALSE, message=FALSE}
findcutnum(
  factor = dat1$invasion,  # Predictor variable (invasion)
  outcome = dat1$LVSI,     # Outcome variable (LVSI)
  datatype = "logistic",   # Data type for logistic regression
  nmin = 8,                # Minimum group size
  segment = 100)           # Number of segments
```

In this dataset, which contains only 30 entries, the number of each
segment is set to 8 (nmin=8), while the number of segments remains at
the default value of 100 (segment=100). The aim is to find the optimal
locations of cutoff points for invasion in response to the presence of
LVSI. Therefore, invasion is the variable factor, and LVSI is the
outcome. As illustrated above, the number of cutoff points needed in
this case is 2.

Now, the function `findcut` will be used to determine the optimal
locations of the two cutoff points.

```{r findcut_2, warning=FALSE, message=FALSE}
findcut(
  factor = dat1$invasion,   # Predictor variable (invasion)
  outcome = dat1$LVSI,      # Outcome variable (LVSI)
  cutnum = 2,              # Number of cutpoints
  datatype = "logistic",   # Data type for logistic regression
  nmin = 8,                # Minimum group size
  segment = 100            # Number of segments
)
```

In this dataset of 30 entries, the number of each segment is set to 8
(nmin=8), while the number of segments remains at the default value of
100 (segment=100). The goal is to find the optimal locations of cutoff
points for invasion in response to the presence of LVSI. Invasion is the
variable factor, and LVSI is the outcome. The required number of cutoff
points in this case is 2.

The best locations of the cutoff points are determined by two methods:

-   **`logtest_likelihood.ratio.test`**: This method tests all possible
    cutoffs using the likelihood ratio test. The cutoff that yields the
    minimal p-value is considered the best location.

-   **`AUC`**: This method tests all possible cutoffs using the AUC
    method. The cutoff that yields the maximal AUC is considered the
    best location.

```{r sessionInfo, warning=FALSE, message=FALSE}
sessionInfo()
```
